Section Four

### 2020-10-06 new2020/sections/sec03.Rmd
Some words for Cdf stuff - and ready for more data/graphs/tables
More results + finally figure out there's a bug in the statsoc class re tables...

## Particle filter ETAs
@label:pf_etas;

@@

## Results
@label:pf_results;

@@

```{r overall_results,echo=FALSE,cache=TRUE,results="asis",dependson="load_results"}
res_overall <- eta_talk %>%
   filter(time_until_arrival > 0) %>%
   summarize(
      pf_rmse = sqrt(mean(pf_error^2)),
      pf_mae = mean(abs(pf_error)),
      pf_mape = mean(abs(pf_error / time_until_arrival)) * 100,
      pf_cicov = mean(pf_ci) * 100,
   #   pf_ci_width = mean(pf_upper - pf_lower),
   #   pf_ci_lower = quantile(pf_upper - pf_lower, 0.025),
   #   pf_ci_upper = quantile(pf_upper - pf_lower, 0.975),
      gtfs_rmse = sqrt(mean(gtfs_error^2)),
      gtfs_mae = mean(abs(gtfs_error)),
      gtfs_mape = mean(abs(gtfs_error / time_until_arrival)) * 100
   ) %>%
   pivot_longer(everything(),
      names_to = c("model", ".value"),
      names_pattern = "([a-z]*)_(.*)"
   ) %>%
   mutate(
      model = c("Particle filter", "Schedule-delay")
   ) %>%
   rename(
      " " = "model",
      "RMSE (s)" = "rmse",
      "MAE (s)" = "mae",
      "MAPE (%)" = "mape",
      "PICP (%)" = "cicov"
   )

opts <- options(knitr.kable.NA = "")
knitr::kable(res_overall, #"latex",
   booktabs = TRUE,
   # table.envir = "table",
   digits = 0,
   caption = "Results for particle filter and other."
)
options(opts)
```

Since traffic and bus numbers change over the course of a day, we also computed the assessment statistics over 15\ minute intervals. These results are shown in @cref:fig:day_results;, and immediately we can see the sizeable affect of peak time of predictive accuracy. This suggests that, during off-peak time when traffic conditions are stable, the real-time predictions are a vast improvement on the status quo. Since the underlying network model is not yet enabled to handle forecasts, its not surprising that accuracy and uncertainty coverage is lacking.

```{r day_results,echo=FALSE,fig.width=8,fig.height=4,message=FALSE,warning=FALSE,cache=TRUE,fig.cap="\\label{fig:day_results}Prediction accuracy results for the particle filter and GTFS methods.",dependson="load_results"}
library(patchwork)

res_timeofday <- eta_talk %>%
    mutate(
        hour = timestamp %>% format("%H") %>% as.integer,
        minute = timestamp %>% format("%M") %>% as.integer,
        hour = hour + (floor(minute / 15) * 15 / 60)
    ) %>%
    filter(time_until_arrival > 0) %>%
    group_by(hour) %>%
    summarize(
        pf_rmse = sqrt(mean(pf_error^2)),
        pf_mae = mean(abs(pf_error)),
        pf_mape = mean(abs(pf_error / time_until_arrival)),
        pf_ci_cov = mean(pf_ci),
        pf_ci_width = mean(pf_upper - pf_lower),
        pf_ci_lower = quantile(pf_upper - pf_lower, 0.025),
        pf_ci_upper = quantile(pf_upper - pf_lower, 0.975),
        gtfs_rmse = sqrt(mean(gtfs_error^2)),
        gtfs_mae = mean(abs(gtfs_error)),
        gtfs_mape = mean(abs(gtfs_error / time_until_arrival))
    ) %>% arrange(hour)

pal <- viridis::magma(3)[2:1] #RColorBrewer::brewer.pal(2, "Set1")
names(pal) <- c("Particle filter", "GTFS")

plot_base <- ggplot(res_timeofday, aes(hour)) +
    theme_classic() +
    theme(legend.position = "none") +
    scale_colour_manual(values = pal, name = "") +
    xlab("Time of day (hour)")

plot_rmse <- plot_base +
    geom_path(aes(y = pf_rmse, colour = "Particle filter")) +
    geom_path(aes(y = gtfs_rmse, colour = "GTFS"), alpha = 0.5) +
    scale_y_continuous(
        "RMSE (s)"
        # trans = "log10"
    ) +
    theme(legend.position = "bottom")

plot_mae <- plot_base +
    geom_path(aes(y = pf_mae, colour = "Particle filter")) +
    geom_path(aes(y = gtfs_mae, colour = "GTFS"), alpha = 0.5) +
    scale_y_continuous(
        "MAE (s)"
        # trans = "log10"
    )

plot_mape <- plot_base +
    geom_path(aes(y = 100 * pf_mape, colour = "Particle filter")) +
    geom_path(aes(y = 100 * gtfs_mape, colour = "GTFS"), alpha = 0.5) +
    scale_y_continuous(
        "MAPE (%)"
        # trans = "log10"
    )

plot_ci <- plot_base +
    geom_hline(aes(yintercept = 85), lty = 2) +
    geom_path(aes(y = 100 * pf_ci_cov, colour = "Particle filter")) +
    scale_y_continuous(
        name = "PICP (%)"
    )

plot_rmse + plot_mae +
    plot_mape + plot_ci +
    guide_area() +
    plot_layout(
        guides = "collect",
        design = "
            13
            24
            55
        ",
        heights = c(1, 1, 0.1)
    )
```

>>>>

## Particle filter ETAs

@@

## Results

@@

```{r overall_results,echo=FALSE,cache=TRUE,results="asis",dependson="load_results"}

res_overall <- eta_talk %>%
   filter(time_until_arrival > 0) %>%
   summarize(
      pf_rmse = sqrt(mean(pf_error^2)),
      pf_mae = mean(abs(pf_error)),
      pf_mape = mean(abs(pf_error / time_until_arrival)) * 100,
      pf_cicov = mean(pf_ci) * 100,
   #   pf_ci_width = mean(pf_upper - pf_lower),
   #   pf_ci_lower = quantile(pf_upper - pf_lower, 0.025),
   #   pf_ci_upper = quantile(pf_upper - pf_lower, 0.975),
      gtfs_rmse = sqrt(mean(gtfs_error^2)),
      gtfs_mae = mean(abs(gtfs_error)),
      gtfs_mape = mean(abs(gtfs_error / time_until_arrival)) * 100
   ) %>%
   pivot_longer(everything(),
      names_to = c("model", ".value"),
      names_pattern = "([a-z]*)_(.*)"
   ) %>%
   mutate(
      model = c("Particle filter", "Schedule-delay")
   ) %>%
   rename(
      " " = "model",
      "RMSE (s)" = "rmse",
      "MAE (s)" = "mae",
      "MAPE (%)" = "mape",
      "PICP (%)" = "cicov"
   )

opts <- options(knitr.kable.NA = "")
knitr::kable(res_overall, #"latex",
   booktabs = TRUE,
   format = "latex",
   # label = "overall_results",
   digits = 0,
   caption = "Results for particle filter and other.",
   caption.short = "Short cap"
)
options(opts)
```

### 2020-10-06 new2020/sections/sec04.Rmd
More results + finally figure out there's a bug in the statsoc class re tables...

---
title: Section Four
header-includes:
   - \usepackage{cleveref}
   - \input{../symbols.tex}
---

@@

Computing a CDF approximation using particles is in fact a trivial exercise, and can be done in real-time without any negative affect on performance. This is because the method described below *does not* require sorting of the particles, which we would need to do to compute, for example, quantiles (median, or a prediction interval). This essentially consists of counting the number of particles in each one-minute interval and expressing it as a probability. Here, the particle arrival time $\alpha\vi$ is in seconds, is transformed to minutes and rounded down using the floor operator, while the discrete CDF is reported in $a$\ minutes.
\begin{equation}
   \label{eq:eta_count}
   \Pr(A \in [a, a+1]) = 
   \frac{1}{\tilde N} \sum_{i=1}^{\tilde N} I\left\{ \lfloor\frac{\alpha\vi}{60}\rfloor = a \right\},
\end{equation}
where the indicator $I\{a=b\}$ is 1 if $a$ equals $b$, and zero otherwise.

@@

```{r pf_cdf,echo=FALSE,fig.width=6,fig.height=4,message=FALSE,warning=FALSE,cache=TRUE,fig.cap="\\label{fig:pf_cdf}Particle filter estimates of arrival time and the process to obtain a discretized CDF approximation."}
library(tidyverse)
set.seed(346)
N <- 2e3
x <- rlnorm(N, log(14), 0.14)
d <- data.frame(x = x)

@@

Of course, we can obtain simple summary statistics for the distribution, such as the median or other quantiles. For a quantile $q\in(0,1)$, we calculate
\[
   \hat A_q = \max\{a\in\{0,1,\ldots\} : \Pr(A < a) \leq q\}
\]
which provides the smallest arrival time, in minutes, with a probability of no more than $q$ of the bus arriving before that time.

@@

Calculating these events is straighforward given the CDF, and can be performed on a user's phone or quickly on the server. We performed a journey planning exercise to assess the reliability of these results by computing the probablity of a bus arriving before a specific time and assessing the outcome. Due to restrains on the current implementation, we only have arrival time distributions for buses which have begun the route. The journey planning problem is this: *given two alternative bus stop options requiring walking in opposite directions, which is the best to take?* We account for the walking time to both stops A and B, and compute the probabilities of catching the next few buses.

@@

## Simplified ETA CDF
@label:eta_cdf;

* round to minutes
* compute the CDF by definition ``number of particles arriving within $x$ minutes''

>>>>

---
title: Section Four
header-includes:
   - \usepackage{cleveref}
   - \usepackage{booktabs}
   - \input{../symbols.tex}
---

```{r,echo=FALSE}
knitr::opts_chunk$set(fig.path="figures/")
```

@@

Computing a CDF approximation using particles is in fact a trivial exercise, and can be done in real-time without any negative affect on performance. This is because the method described below *does not* require sorting of the particles, which we would need to do to compute, for example, quantiles (median, or a prediction interval). This essentially consists of counting the number of particles in each one-minute interval and expressing it as a probability. Here, the particle arrival time $\alpha\vi$ is in seconds, is transformed to minutes and rounded down using the floor operator, while the discrete CDF is reported in $a$\ minutes.
\begin{equation}
   \label{eq:eta_count}
   \Pr{A \in [a, a+1]} =
   \frac{1}{\tilde N} \sum_{i=1}^{\tilde N} I\left\{ \lfloor\frac{\alpha\vi}{60}\rfloor = a \right\},
\end{equation}
where the indicator $I\{a=b\}$ is 1 if $a$ equals $b$, and zero otherwise.

@@

```{r pf_cdf,echo=FALSE,fig.width=6,fig.height=4,out.width="\\textwidth",message=FALSE,warning=FALSE,cache=TRUE,fig.cap="Particle filter estimates of arrival time and the process to obtain a discretized CDF approximation."}
library(tidyverse)
set.seed(346)
N <- 2e3
x <- rlnorm(N, log(14), 0.14)
d <- data.frame(x = x)

@@

Of course, we can obtain simple summary statistics for the distribution, such as the median or other quantiles. For a quantile $q\in(0,1)$, we calculate
\[
   \hat A_q = \max\{a\in\{0,1,\ldots\} : \Pr{A < a} \leq q\}
\]
which provides the smallest arrival time, in minutes, with a probability of no more than $q$ of the bus arriving before that time.

@@

Calculating these events is straighforward given the CDF, and can be performed on a user's phone or quickly on the server. We performed a journey planning exercise to assess the reliability of these results by computing the probablity of a bus arriving before a specific time and assessing the outcome. Due to restrains on the current implementation, we only have arrival time distributions for buses which have begun the route. The journey planning problem is this: *given two alternative bus stop options requiring walking in opposite directions, which is the best to take?* We account for the walking time $W_1$ and $W_2$ to both stops $R_1$ and $R_2$, respectively, and compute the probabilities of catching the next few buses. The probability of catching $R_j$, $j=1,2$, if leaving now at time $T$, is given by
\[
   \Pr{\text{catch} | \text{choose } R_j} =
   \Pr{A_j \geq T + W_j} =
   1 - \Pr{A_j < T + W_j}
\]
which is easily obtained from the CDF. To simulate, we chose a sequence of times from 9\ am til 3\ pm in 15\ minute intervals, and for each calculated the probability of catching upcoming buses at the two alternative stops. This was compared to the outcome, and the GTFS binary prediction. The results are shown in @cref:fig:jp_catch;.

```{r jp_prep,echo=FALSE,message=FALSE,cache=TRUE}
library(tidyverse)
library(patchwork)
load("../data/all_results.rda")
```

```{r jp_catch,cache=TRUE,dependson="jp_prep",echo=FALSE,fig.width=8,fig.height=2.5,out.width="\\textwidth",message=FALSE,warning=FALSE,fig.cap="Predicted outcomes of journey planning process."}
## Now plot p-catch versus did_catch
mdata <- map(all_results, "summary") %>%
  bind_rows %>%
  mutate(
    did_catch = as.integer(did_catch),
    did_arrive = as.integer(did_arrive),
    gtfs_catch = ifelse(gtfs_catch == 1, "Yes", "No"),
    gtfs_arrive = ifelse(gtfs_arrive == 1, "Yes", "No")
  )

mdata %>%
   mutate(did_catch = ifelse(did_catch, "Bus was caught", "Bus was missed")) %>%
   ggplot(aes(p_catch + rnorm(length(p_catch), 0, 0.01), gtfs_catch)) +
      ggbeeswarm::geom_beeswarm(
         groupOnX = FALSE,
         priority = "random"
      ) +
      theme_classic() +
      scale_x_continuous("Predicted probabilty of catching the bus") +
      scale_y_discrete("GTFS prediction") +
      facet_grid(~did_catch) +
      theme(
         strip.background = element_blank(),
         strip.text = element_text(hjust = 0)
      )

gtfs_catch_tab <- table(
  gtfs = mdata$gtfs_catch,
  outcome = ifelse(mdata$did_catch, "Yes", "No")
)
gtfs_catch_pc <-
  round(sweep(gtfs_catch_tab, 1, rowSums(gtfs_catch_tab), "/"), 2)

pf_catch_tab50 <- table(
   pf = ifelse(mdata$p_catch > 0.5, "Yes", "No"),
   outcome = ifelse(mdata$did_catch, "Yes", "No")
)
pf_catch_pc50 <-
   round(sweep(pf_catch_tab50, 1, rowSums(pf_catch_tab50), "/"), 2)

pf_catch_tab80 <- table(
   pf = ifelse(mdata$p_catch > 0.8, "Yes", "No"),
   outcome = ifelse(mdata$did_catch, "Yes", "No")
)
pf_catch_pc80 <-
   round(sweep(pf_catch_tab80, 1, rowSums(pf_catch_tab80), "/"), 2)

tab <- cbind(
  "Predicted outcome" = c("No", "Yes"),
  format(gtfs_catch_pc, nsmall = 2L, digits = 2L),
  NA,
  format(pf_catch_pc50, nsmall = 2L, digits = 2L),
  NA,
  format(pf_catch_pc80, nsmall = 2L, digits = 2L)
)

# roc curve?
pc <- seq(0, 1, by = 0.05)
roc <- lapply(pc,
   function(p) {
      tab <- table(
         pf = factor(ifelse(mdata$p_catch > p, "Yes", "No"), levels = c("No", "Yes")),
         outcome = factor(ifelse(mdata$did_catch, "Yes", "No"), levels = c("No", "Yes"))
      )
      P <- sum(tab[,2])
      N <- sum(tab[,1])
      TP <- tab[2,2]
      TN <- tab[1,1]
      FP <- tab[2,1]
      FN <- tab[1,2]
      c(
         p = p,
         tpr = TP / P,
         tnr = TN / N,
         ppv = TP / (TP + FP),
         fnr = FN / P,
         fpr = FP / N
      )
   }
) %>% bind_rows

gtfs_roc <- data.frame(
   fpr = gtfs_catch_tab[2,1] / sum(gtfs_catch_tab[,1]),
   tpr = gtfs_catch_tab[2,2] / sum(gtfs_catch_tab[,2])
)
```

```{r jp_catch_dec,cache=TRUE,dependson="jp_catch",echo=FALSE,fig.width=5,fig.height=2.5,out.width="0.5\\textwidth",fig.align="center",message=FALSE,warning=FALSE,fig.cap="False positive rate for varying decision level. Increasing the threshold for predicting a positive outcome reduces the *miss rate*, at the cost of not trying to catch potential buses."}
gp <- 0.175
ggplot(roc, aes(p, fpr)) +
   geom_path(lwd = 1.2) +
   coord_cartesian(expand = FALSE) +
   geom_segment(aes(x = gp, xend = 0, yend = fpr),
      data = gtfs_roc, colour = "red", lty = 2
   ) +
   geom_segment(aes(x = gp, xend = gp, yend = 0),
      data = gtfs_roc, colour = "red", lty = 2) +
   xlab("Decision threshold") +
   ylab("False positive / miss rate") +
   theme_classic()
```

```{r jp_catch_tab,cache=TRUE,dependson="jp_catch",echo=FALSE,results='asis'}
options(knitr.kable.NA = "")
knitr::kable(
  tab,
  row.names = FALSE,
  format = "latex",
  booktabs = TRUE,
#   label = "pf_catch_tab",
  caption = "Prediction results for all journeys, with the displayed values representing the proportion of outcomes in each cell (rows are conditioned by the prediction).",
  caption.short = "Prediction results for all journeys",
  escape = FALSE,
  linesep = ""
) %>%
  kableExtra::add_header_above(
    c(" ", "Schedule-delay" = 2, " ", "P > 0.5" = 2, " ", "P > 0.8" = 2)
  ) %>%
  kableExtra::add_header_above(
    c("", "Observed outcome" = 8)
  ) %>%
  kableExtra::kable_styling(font_size = 8)
```

Discussion of results!!

* we can adjust decision to fit the situation - coffee with friend vs job interview
* doesn't account for *wait time* (i.e., if you miss the bus, how long till the next one)

The second journey planning scenario is *arriving before a specified time*, which requires forecasting further into the future. Such a decision might be important for someone, allowing them to either inform their friend they'll be late for that coffee, or catching a taxi to get to the interview. As before, we selected a stop and this time asked if the bus would arrive on-time (where on time was 15--45\ minutes in the future). The results, as before, are displayed in, where the successfull event is now that the bus arrived before the specified time $A$.

```{r jp_arrive,cache=TRUE,dependson="jp_catch",echo=FALSE,fig.width=8,fig.height=2.5,out.width="\\textwidth",message=FALSE,warning=FALSE,fig.cap="Predicted outcomes of journey planning process - arrive on time."}
## Now plot p-catch versus did_catch

mdata %>%
   mutate(did_arrive = ifelse(did_arrive, "Arrived on time", "Arrived late")) %>%
   ggplot(aes(p_arrive + rnorm(length(p_arrive), 0, 0.01), gtfs_arrive)) +
      ggbeeswarm::geom_beeswarm(
         groupOnX = FALSE,
         priority = "random"
      ) +
      theme_classic() +
      scale_x_continuous("Predicted probabilty of bus arriving on time") +
      scale_y_discrete("GTFS prediction") +
      facet_grid(~did_arrive) +
      theme(
         strip.background = element_blank(),
         strip.text = element_text(hjust = 0)
      )

gtfs_arrive_tab <- table(
  gtfs = mdata$gtfs_arrive,
  outcome = ifelse(mdata$did_arrive, "Yes", "No")
)
gtfs_arrive_pc <-
  round(sweep(gtfs_arrive_tab, 1, rowSums(gtfs_arrive_tab), "/"), 2)

pf_arrive_tab50 <- table(
   pf = ifelse(mdata$p_arrive > 0.5, "Yes", "No"),
   outcome = ifelse(mdata$did_arrive, "Yes", "No")
)
pf_arrive_pc50 <-
   round(sweep(pf_arrive_tab50, 1, rowSums(pf_arrive_tab50), "/"), 2)

pf_arrive_tab80 <- table(
   pf = ifelse(mdata$p_arrive > 0.8, "Yes", "No"),
   outcome = ifelse(mdata$did_arrive, "Yes", "No")
)
pf_arrive_pc80 <-
   round(sweep(pf_arrive_tab80, 1, rowSums(pf_arrive_tab80), "/"), 2)

tab <- cbind(
  "Predicted outcome" = c("No", "Yes"),
  format(gtfs_arrive_pc, nsmall = 2L, digits = 2L),
  NA,
  format(pf_arrive_pc50, nsmall = 2L, digits = 2L),
  NA,
  format(pf_arrive_pc80, nsmall = 2L, digits = 2L)
)

# roc curve?
pc <- seq(0, 1, by = 0.05)
roc <- lapply(pc,
   function(p) {
      tab <- table(
         pf = factor(ifelse(mdata$p_arrive > p, "Yes", "No"), levels = c("No", "Yes")),
         outcome = factor(ifelse(mdata$did_arrive, "Yes", "No"), levels = c("No", "Yes"))
      )
      P <- sum(tab[,2])
      N <- sum(tab[,1])
      TP <- tab[2,2]
      TN <- tab[1,1]
      FP <- tab[2,1]
      FN <- tab[1,2]
      c(
         p = p,
         tpr = TP / P,
         tnr = TN / N,
         ppv = TP / (TP + FP),
         fnr = FN / P,
         fpr = FP / N
      )
   }
) %>% bind_rows

gtfs_roc <- data.frame(
   fpr = gtfs_arrive_tab[2,1] / sum(gtfs_arrive_tab[,1]),
   tpr = gtfs_arrive_tab[2,2] / sum(gtfs_arrive_tab[,2])
)
```

```{r jp_arrive_dec,cache=TRUE,dependson="jp_arrive",echo=FALSE,fig.width=5,fig.height=2.5,out.width="0.5\\textwidth",fig.align="center",message=FALSE,warning=FALSE,fig.cap="False positive rate for varying decision level. Increasing the threshold for predicting a positive outcome reduces the arrival late rate, at the cost of not trying to catch potential buses."}
gp <- 0.175
ggplot(roc, aes(p, fpr)) +
   geom_path(lwd = 1.2) +
   coord_cartesian(expand = FALSE) +
   # geom_segment(aes(x = gp, xend = 0, yend = fpr),
   #    data = gtfs_roc, colour = "red", lty = 2
   # ) +
   # geom_segment(aes(x = gp, xend = gp, yend = 0),
   #    data = gtfs_roc, colour = "red", lty = 2) +
   xlab("Decision threshold") +
   ylab("False positive") +
   theme_classic()
```

```{r jp_arrive_tab,cache=TRUE,dependson="jp_arrive",echo=FALSE,results='asis'}
options(knitr.kable.NA = "")
knitr::kable(
  tab,
  row.names = FALSE,
  format = "latex",
  booktabs = TRUE,
#   label = "pf_catch_tab",
  caption = "Prediction results for all journeys 2, with the displayed values representing the proportion of outcomes in each cell (rows are conditioned by the prediction).",
  caption.short = "Prediction results for all journeys 2",
  escape = FALSE,
  linesep = ""
) %>%
  kableExtra::add_header_above(
    c(" ", "Schedule-delay" = 2, " ", "P > 0.5" = 2, " ", "P > 0.8" = 2)
  ) %>%
  kableExtra::add_header_above(
    c("", "Observed outcome" = 8)
  ) %>%
  kableExtra::kable_styling(font_size = 8)
```

The final scenario is for a journey which requires transfering between two services, in which case we want to predict the probability that a transfer between two particular trips at the transfer stop will be successful (that is, the first bus arrives before the second).

@@
